A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2014-03-04 response/response.tex

  \end{itemize}}

  \item
Pg. 26: Please clarify whether storage is needed in automatic differentiation. For
example, do we need to store w, v , and u?

{\color{red}  
\begin{itemize}
     \item
      Clarification is added in the manuscript.  
  \end{itemize}}

\subsection*{Typos/editing suggestions}
\begin{enumerate}

\item Pg. 1: In line 43, remove “the” in “all the of the code”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 1: In line 46, replace “numeric differentiation” by “numerical differentiation”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 2: In line 25, the sentence “The distinction of “accurate” is deﬁned by ...”
awkward. Please revise it.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 2: In lines 32 and 48, “ﬁnite-difference” should be “forward-difference”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 2: In lines 33, I believe “aide” should be “aid”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 3: In line 12, I would replace “such as required” by “such as those required”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 3: In line 22, replace “with each the methods” by “with each of the methods”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 3: In line 25, replace “implementing complex-step” by “implementing the complex-
step method”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 3: I would replace the title of Section 1.1, “Differentiation Techniques” by “Dif-
ferentiation techniques” for consistency, i.e., only capitalize the ﬁrst letter of the ﬁrst
word. Similarly, for the rest of the section titles.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 4: In line 10, “be be made complex” should be “be made complex”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 4: In Eq. (1), replace comma by period.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 4: In line 22, you have an extra white space in “O (h2 )”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 4: I would check the proper use of hyphens along the manuscript. For example,
in the title of Section 1.1.2, I would replace “Automatic-differentiation” by “Automatic
differentiation” (i.e., no hyphen).

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 4: In line 42, replace “based the chain-rule” by “based on the chain rule”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 4: In lines 45-46, “(add, multiply, power, transcendental and the like)” should
probably be “(addition, multiplication, exponentiation, and the like)” or similar.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 4: In line 47, I would replace “used in the study” by “used in this study”. Similarly,
in other instances of the manuscript.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 5: In line 39, you may replace “the par tials” by “the par tial derivatives”. Similarly
in page 26.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 5: I would replace the title of Section 1.2 “Tangent-stiffness” by “Tangent-stiffness
matrix”. Similarly, in line 50.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 6: In line 14, “quasi-Newton’s method’s” should probably be “quasi-Newton
methods”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 6: In line 18, “about about a par ticular” should be “about a par ticular”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 6: In line 51, you may replace “argument to F ” by “argument of F ”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 7: In line 11, you may replace “single independent vector component of the
function” by “single independent vector component of the argument of the function”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 7: In line 24, replace “Where Kij is” by “where Kij is”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 7: In line 40, replace “in literature [19]” by “in the literature [19]”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 8: Remove the period at the end of the titles of Sections 2 and 2.1, for consis-
tency. Similarly, for the rest of the manuscript. Also, use italic font for “Peridigm” in
the title of §2.1.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 9: In line 29, replace “Perdigm” by “Peridigm”. Similarly in page 13 (line 44).

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 9: In line 37, replace “be simply be declared” by “be simply declared”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 9: In line 55, I would replace “computational scientists” by “computer scientists”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 10: In line 30, I would replace “The purpose of applying the load” by “Applying
the load”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 13: I would add a comma at the end of Eq. (8), and replace “Where D is
distance” by “where D is distance” in line 14.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 13: In line 15, replace “based method, M is” by “based method, and M is”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 13: In line 31, replace “correspond to” by “corresponds to”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 14: In lines 19-21, you have twice the word “nonzero” in “number of nonzero
tangent-stiffness (TS) non-zero matrix elements”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 14: In lines 22, I would replace “The units for accuracy are derived equation (8)
and the units used to deﬁne bulk material proper ties, mentioned in Section 2.2.” by
“The accuracy measure is given by equation (8) and the choice of elastic moduli is
mentioned in Section 2.2.”

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 15: In line 54, it seems that “yet still fall” should be “yet still falls”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 17: In line 45 you say “h, cannot be too small,”.
I understand that you mean
that you can take h as small as possible. However, the way written may have the
opposite connotation. I would revise it for clarity.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 18: In lines 35-36, replace “the number nonzero” by “the number of nonzero”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 18: In line 39, replace “the the number of nonzero” by “the number of nonzero”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 18: In line 50, I would replace “experimental results,” by “computational experi-
ments,”. Otherwise, it may give the impression of physical experiments.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 20: In line 13, is the use of the word “order” in “number of iterations taken order
to measure” correct?

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 20: In lines 26-27, replace “complex-step all of the methods” by “complex-step
in all of the methods”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 20: In line 50, you may either replace “to implement; It only” by “to implement;
it only”; or replace the semicolon by a period.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pgs. 21-23: You may capitalize the ﬁrst letters on the journal names in your refer-
ences [9, 20]

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 22: In your reference [10], you may replace “fr ´echet” by “Fr ´echet”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 23: Update your reference [21] (see the reference [3] in this repor t). Also,
update your reference [27].

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 24: In line 18, replace the comma by a period.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 24: I would replace the comma by a period in line 28, and replace “using
L’H ˆopital’s rule” by “Using l’H ˆopital’s rule” in line 31.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 24: I would replace the comma by a period in line 34, and replace “of course,
as h → 0” by “As h → 0” in line 37.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 24: I would replace the title of Appendix B “AD Example” by “Automatic differ-
entiation example”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 24: In line 45, you may replace “sin(cos(x))” by “sin(cos(x))”. For that, use \sin
instead of sin, etc. (similarly in the bottom of the page and in page 25). Also, add
a period at the end of the expression.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pg. 25: In line 55, you may remove the “·” symbol in “ d
dx · g(h(k(x)))”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

 \item Pgs. 25 and 26: I would number the table in 2(c) (page 25). Let’s assume the
number is “2”. Then, in line 13 (page 26), I would replace “chosen for each function
and held as constant” by “chosen for each function as in Table 2 and held constant”.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

\item Pg. 26: In line 24, add a period at the end of the expression.

{\color{red}  
\begin{itemize}
     \item
      apples
  \end{itemize}}

### 2014-03-04 jacobian_calc_CMAME.tex
typos

In order to maintain the quadratic convergence properties of Newton's method in
quasi-static nonlinear analysis of solid structures it is crucial to obtain
accurate, algorithmically consistent tangent-stiffness matrices. A goal of the
study described in this paper was to establish the suitability of an
under-explored method for numerical computation of tangent-stiffness operators,
referred to as ``complex-step'', and compare the \note[Michael]{re: I.1}
\remove[Michael]{new} method with other techniques for numerical derivative
calculation: automatic differentiation, forward finite-difference, and central
finite-difference. The complex-step method was \note[Michael]{re: I.1, could
suggest claim of novelty} \remove[Michael]{newly} implemented in a massively
parallel computational peridynamics code for the purpose of this comparison.
The methods were compared through profiling of the code for accuracy,
speed, efficiency, and parallel scalability. The research provides data that
can serve as practical guide for code developers and analysts faced with
choosing which method best suits the needs of their application code.
Additionally, motivated by the reproducible research movement, all the of the
code, examples, and workflow to regenerate the data and figures in this paper
are provided as open source.

@@

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
Newton's method \sep Newton-Raphson \sep numeric differentiation \sep complex-step \sep finite-difference \sep automatic-differentiation \sep Jacobian \sep tangent-stiffness
%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)
%\MSC [2010] 65D25 
\end{keyword}

@@

A goal of this study was to develop and evaluate a \note[Michael]{re: I.1}
\change[Michael]{new}{relatively underexplored}, accurate, and practical method
for calculating tangent-stiffness matrices against established methods.  This
new method is based on a complex number Taylor series expansion and referred to
as CTSE or the ``complex-step'' method.  The distinction of ``accurate'' is
defined by comparison of the \note[Michael]{re: I.1}\remove[Michael]{new}
method to popular finite differencing techniques used for computing
tangent-stiffness matrices and with the exact to machine precision
algorithmically consistent derivatives computed with \emph{automatic
differentiation}. \note[Michael]{re: II.1}For the reader's convenience, 
from plasticity theory, the algorithmic consistency of
the tangent-stiffness matrix refers to consistency with the integration of a
plasticity model under the Khun-Tucker constraints or similar where the tangent-stiffness matrix is
computed using the elastoplastic modulii that are determined from the integration \cite{simo1998}.

Comparative data regarding the accuracy and compute cycle timing for the
complex step, finite-difference, central-difference, and automatic
differentiation methods are included to aide developers and analysts in
computational mechanics code design who are faced with implementing a general
method for tangent stiffness matrix calculation.  A comparison of the methods was achieved through instrumentation of a
massively parallel computational mechanics code.

The scope of the application component of the study was limited to a single
material model, an elastic peridynamic solid, implemented in the computational
peridynamics code, \textit{Peridigm} \cite{peridigm}. Identifying a specific
application provided a practical framework for implementing the
\note[Michael]{re: I.1}\remove[Michael]{new} method and solving engineering
problems to generate the data needed to compare the methods. In particular,
\emph{Peridigm} was chosen because it combined several helpful
characteristics: the utilization of Newton's method and tangent-stiffness
matrices for solving nonlinear quasi-static problems, prior inclusion of
finite-difference, central-difference, and automatic-differentiation
methods needed for comparison to the new complex-step method
implementation, and being an agile-components code that makes use of
distributed computing data structures via Trilinos \cite{trilinos} for
efficient parallelization on large clusters. Agile-components is a software
development term meaning to structure a software project for continuous
development to respond to evolving specifications by making much use of
techniques from object-oriented programming and reusing proven pre-existing
code to reduce development time and increase reliability. The modularity of
\emph{Peridigm} allowed the adding of new features, such as required by the
study.

The aim of this paper is not only to \note[Michael]{re: I.1}\change[Michael]{introduce the new}{study the} complex-step method \change[Michael]{in
the context of}{of} evaluating tangent-stiffness matrices, but to serve as a review
for other differentiation techniques useful for solving non-linear systems with
Newton's method. After presenting background information on the underlying
methods to be discussed, presented in this paper are: In subsection~\ref{subsec:TS} a description of
tangent-stiffness matrices, detailed directions for producing
tangent-stiffness matrices with each of the methods identified above, and a
description and justification of the \note[Michael]{re: I.1}\remove[Michael]{new}complex-step method for calculating
tangent-stiffness matrices, in sub-subsection~\ref{subsubsec:Impl} a description of implementing complex-step in
the \emph{Peridigm} software, in section~\ref{JGAM} a description of the quantities-of-interest
used to rank the methods, in section~\ref{sec:Res} a presentation and analysis of the results of the
comparative study, and in section~\ref{sec:Conc} conclusions and thoughts on potential future work.
Finally, in the interest of replicable research, the reader is referred to the
corresponding author's website for \emph{Peridigm} C++ source-code and data
necessary to reproduce the results presented here. Additionally, included on
the website is an example serial C++ library with classes for solving
non-linear systems using the techniques discussed here, as well as two
validation and verification example problems which show how to use the library.
The purpose of the serial C++ library is to demonstrate in a less complex
manner than in the \emph{Peridigm} code how one would use
automatic-differentiation or complex-step to better encourage their use and
discussion. \note[Michael]{re: II.5}This open source resource is available at the corresponding author's website: http://idl.utsa.edu/jtfoster/paper-repositories/
or directly at https://github.com/utsa-idl/Complex_Step_Jacobian_Project_Archive.

\subsection{Differentiation Techniques}

This section contains background information on the differentiation techniques
underlying the tangent-stiffness matrix calculation methods compared in the
study. Since the derivation of first-order finite-difference techniques from
Taylor series expansion is considered well known, it is not described here.
Instead the reader is referred to \cite[Chap. 4.1.3]{chapra2010}.

\subsubsection{The ``complex-step'' method.}  \label{sec:CSmethod} It is
possible to approximate derivatives quite accurately with a technique based on
a complex variable Taylor series expansion of a function.  This method was
first described by Lyness and Moler
\cite{lyness1967numerical,lyness1968differentiation} and has more recently been
rediscovered for use in engineering analysis.  The basic idea is that a model parameter can
be be made complex and expanded in a Taylor series about a small perturbation
along the imaginary axis by some arbitrary value $h$ as follows
%
\begin{align} f ( x + i h ) =& f(x) + \frac{\partial f(x)}{\partial x} \frac{i
h}{1!}  \notag \\ &+ \frac{\partial^2 f(x)}{\partial x^2} \frac{(i h)^2}{2!} +
\frac{\partial^3 f(x)}{\partial x^3} \frac{(i h)^3}{3!} + \cdots,
\label{eqn:complexTaylor} \end{align}
%
Taking the imaginary part of both sides of equation (\ref{eqn:complexTaylor})
and solving for the first derivative and ignoring terms of $\mathcal{O}\left (
h^2 \right)$ yields an estimate
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{\mbox{Im}
\left( f (x + i h) \right)}{h} +\mathcal{O}\left ( h^2 \right).  \label{eqn:complexFirstDeriv} \end{equation}
%
Thus, an estimate of the first derivative of a function can be made by only
utilizing one functional evaluation of a perturbed model parameter along the
imaginary axis.  The step-size $h$ is arbitrary and can be made as small as
practical (even to machine precision without the dangers of roundoff error as
in finite-differences) to yield an accurate estimate of this derivative.  The
only disadvantage of this technique is the necessity of requiring the functions
to accept complex numbers as arguments. Appendix~\ref{sec:appendixA} includes a
simple example that illustrates the use of the complex-step method for the
derivative calculation of a function. 
%
\note[Michael]{re: II.6}
For the reader's convenience the forward difference equivalent to~\ref{eqn:complexFirstDeriv} is
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x)}{h} +\mathcal{O}\left ( h \right). \label{eqn:forwardFirstDeriv} \end{equation},
%
while the central difference equivalent to~\ref{eqn:complexFirstDeriv} is,
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x - h)}{2h} +\mathcal{O}\left ( h^2 \right). \label{eqn:centeredFirstDeriv} \end{equation} \cite[Chap. 4.1.3]{chapra2010}.
%
Note that the order of the truncation error in~\ref{eqn:forwardFirstDeriv} indicates that for a given $h$ forward difference is less accurate an approximation than central difference or complex step.

@@

\subsubsection{Automatic-differentiation} 
\label{ADsubsection}

Automatic-differentiation (AD) is a computerized method for computing exact
derivatives based the chain-rule from calculus. AD takes advantage of the fact
that any mathematical function executed on a computer, no matter how
complicated, is a ``composition of simple operations'' (add, multiply, power,
transcendental and the like) each having known analytical derivatives
\cite{ref-sacado-presentation}. For reference, the AD implementation used in
the study is part of the \emph{Sacado} package from the \emph{Trilinos}
agile-component libraries developed at Sandia National Laboratories
\cite{ref-Sacado}.

An AD system evaluates composition functions in the expected manner: it works
by first evaluating the innermost function of the composition, then presenting
that function's output as input to the next level function until all levels are
complete, observing commonly expected order of operations. However, an AD
system does additional work during a function evaluation in that as each nested
function is evaluated, the function's partial derivative with respect to the
designated variables of the given input is also calculated.  This is possible
because the elementary math functions are hard-coded into the AD source-code
along with their analytical derivatives, and linked by special instructions, so
that when the elementary math functions are called upon for computation, their
partial derivatives may be computed and stored in a sequence. The AD system
then multiplies the final sequence of partial derivatives together to produce
the exact equivalent to taking a partial derivative of the corresponding
composition function with respect to a designated variable at a particular
value. 
%
It is obvious, but bears mentioning, that the AD system could simply
store one value for partial derivatives, modifying it as appropriate for every
function evaluation rather than keeping a sequence. This is significant because
modifying a single value rather than keeping a list of values corresponding to
every level of a composition function, which itself may not be the sole one
needing to be evaluated, represents a savings in memory usage, which at large
scales is an active concern. In the literature, the AD scheme described here is
called \emph{forward automatic-differentiation}. Only forward AD will be
covered here since it is the implementation used in \emph{Sacado} and therefore
in this study; however, the reader is referred to the introduction section of
\cite{ref-AD-methods} and its citations for further information on AD, and
particularly \cite{ref-on-AD} which is foundational.
Appendix~\ref{sec:appendixB} includes an example that walks the reader through
the process a computer uses to compute derivatives via AD.  Some things to note
about AD are that no approximation of derivatives is being made because the
analytical forms of the partials of the elementary math functions are defined
alongside them. The accuracy of AD is then limited by the precision of the AD
system's definition of the elementary math functions and their partial
derivatives.

\subsection{Tangent-stiffness} 
\label{subsec:TS}
In solid and structural computational mechanics the tangent-stiffness matrix is
a linearization operator that describes the stiffness of a system in response
to small displacements imposed upon the current configuration of the system.
Mathematically speaking, the tangent-stiffness represents the gradient of a
high-dimensional energy surface that ``points'' in the minimum direction. While
the terminology of solid mechanics is used here, it's important to note that
these operators appear in other physical settings and are known by other names,
e.g. a \emph{transmissibility matrix} in the context of a Poisson problem or,
more generally, a \emph{Jacobian matrix} in mathematical optimization.

>>>>

In order to maintain the quadratic convergence properties of Newton's method in
quasi-static nonlinear analysis of solid structures it is crucial to obtain
accurate, algorithmically consistent tangent-stiffness matrices. A goal of the
study described in this paper was to establish the suitability of an
underexplored method for numerical computation of tangent-stiffness operators,
referred to as ``complex-step'', and compare the \note[Michael]{re: I.1}
\remove[Michael]{new} method with other techniques for numerical derivative
calculation: automatic differentiation, forward finite difference, and central
finite difference. The complex-step method was \note[Michael]{re: I.1, could
suggest claim of novelty} \remove[Michael]{newly} implemented in a massively
parallel computational peridynamics code for the purpose of this comparison.
The methods were compared through profiling of the code for accuracy,
speed, efficiency, and parallel scalability. The research provides data that
can serve as practical guide for code developers and analysts faced with
choosing which method best suits the needs of their application code.
Additionally, motivated by the reproducible research movement, all of the
code, examples, and workflow to regenerate the data and figures in this paper
are provided as open source.

@@

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
Newton's method \sep Newton-Raphson \sep numerical differentiation \sep complex-step \sep finite difference \sep automatic differentiation \sep Jacobian \sep tangent-stiffness
%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)
%\MSC [2010] 65D25 
\end{keyword}

@@

A goal of this study was to develop and evaluate a \note[Michael]{re: I.1}
\change[Michael]{new}{relatively underexplored}, accurate, and practical method
for calculating tangent-stiffness matrices against established methods.  This
new method is based on a complex number Taylor series expansion and referred to
as CTSE or the ``complex-step'' method. The practical accuracy of CTSE is
determined by comparing its measured accuracy\note[Michael]{re:
I.1}\remove[Michael]{new} relative to the measured accuracy of popular finite
differencing techniques used for computing tangent-stiffness matrices. For this
comparison the exact to machine precision derivatives computed with
\emph{automatic differentiation} serve as the standard. \note[Michael]{re:
II.1}For the reader's convenience, from plasticity theory, the algorithmic
consistency of the tangent-stiffness matrix refers to consistency with the
integration of a plasticity model under the Khun-Tucker constraints or similar
where the tangent-stiffness matrix is computed using the elastoplastic modulii
that are determined from the integration \cite{simo1998}.

Comparative data regarding the accuracy and compute cycle timing for the
complex step, forward difference, central difference, and automatic
differentiation methods are included to aid developers and analysts in
computational mechanics code design who are faced with implementing a general
method for tangent-stiffness matrix calculation.  A comparison of the methods was achieved through instrumentation of a
massively parallel computational mechanics code.

The scope of the application component of this study was limited to a single
material model, an elastic peridynamic solid, implemented in the computational
peridynamics code, \textit{Peridigm} \cite{peridigm}. Identifying a specific
application provided a practical framework for implementing the
\note[Michael]{re: I.1}\remove[Michael]{new} method and solving engineering
problems to generate the data needed to compare the methods. In particular,
\emph{Peridigm} was chosen because it combined several helpful
characteristics: the utilization of Newton's method and tangent-stiffness
matrices for solving nonlinear quasi-static problems, prior inclusion of
forward difference, central difference, and automatic differentiation
methods needed for comparison to the new complex-step method
implementation, and being an agile-components code that makes use of
distributed computing data structures via Trilinos \cite{trilinos} for
efficient parallelization on large clusters. Agile-components is a software
development term meaning to structure a software project for continuous
development to respond to evolving specifications by making much use of
techniques from object-oriented programming and reusing proven pre-existing
code to reduce development time and increase reliability. The modularity of
\emph{Peridigm} allowed the adding of new features, such as those required by the
study.

The aim of this paper is not only to \note[Michael]{re: I.1}\change[Michael]{introduce the new}{study the} complex-step method \change[Michael]{in
the context of}{of} evaluating tangent-stiffness matrices, but to serve as a review
for other differentiation techniques useful for solving non-linear systems with
Newton's method. After presenting background information on the underlying
methods to be discussed, presented in this paper are: In subsection~\ref{subsec:TS} a description of
tangent-stiffness matrices, detailed directions for producing
tangent-stiffness matrices with each of the methods identified above, and a
description and justification of the \note[Michael]{re: I.1}\remove[Michael]{new}complex-step method for calculating
tangent-stiffness matrices, in sub-subsection~\ref{subsubsec:Impl} a description of implementing the complex-step method in
the \emph{Peridigm} software, in section~\ref{JGAM} a description of the quantities-of-interest
used to rank the methods, in section~\ref{sec:Res} a presentation and analysis of the results of the
comparative study, and in section~\ref{sec:Conc} conclusions and thoughts on potential future work.
Finally, in the interest of replicable research, the reader is referred to the
corresponding author's website for \emph{Peridigm} C++ source-code and data
necessary to reproduce the results presented here. Additionally, included on
the website is an example serial C++ library with classes for solving
non-linear systems using the techniques discussed here, as well as two
validation and verification example problems which show how to use the library.
The purpose of the serial C++ library is to demonstrate in a less complex
manner than in the \emph{Peridigm} code how one would use
automatic differentiation or complex-step to better encourage their use and
discussion. \note[Michael]{re: II.5}This open source resource is available at the corresponding author's website: http://idl.utsa.edu/jtfoster/paper-repositories/
or directly at https://github.com/utsa-idl/Complex_Step_Jacobian_Project_Archive.

\subsection{Differentiation techniques}

This section contains background information on the differentiation techniques
underlying the tangent-stiffness matrix calculation methods compared in the
study. Since the derivation of first-order finite difference techniques from
Taylor series expansion is considered well known, it is not described here.
Instead the reader is referred to \cite[Chap. 4.1.3]{chapra2010}.

\subsubsection{The ``complex-step'' method}  
\label{sec:CSmethod}

It is
possible to approximate derivatives quite accurately with a technique based on
a complex variable Taylor series expansion of a function.  This method was
first described by Lyness and Moler
\cite{lyness1967numerical,lyness1968differentiation} and has more recently been
rediscovered for use in engineering analysis.  The basic idea is that a model parameter can
be made complex and expanded in a Taylor series about a small perturbation
along the imaginary axis by some arbitrary value $h$ as follows
%
\begin{align} f ( x + i h ) =& f(x) + \frac{\partial f(x)}{\partial x} \frac{i
h}{1!}  \notag \\ &+ \frac{\partial^2 f(x)}{\partial x^2} \frac{(i h)^2}{2!} +
\frac{\partial^3 f(x)}{\partial x^3} \frac{(i h)^3}{3!} + \cdots.
\label{eqn:complexTaylor} \end{align}
%
Taking the imaginary part of both sides of equation (\ref{eqn:complexTaylor})
and solving for the first derivative and ignoring terms of $\mathcal{O}\left(
h^2 \right)$ yields an estimate
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{\mbox{Im}
\left( f (x + i h) \right)}{h} +\mathcal{O}\left( h^2 \right).  \label{eqn:complexFirstDeriv} \end{equation}
%
Thus, an estimate of the first derivative of a function can be made by only
utilizing one functional evaluation of a perturbed model parameter along the
imaginary axis.  The step-size $h$ is arbitrary and can be made as small as
practical (even to machine precision without the dangers of roundoff error as
in finite differences) to yield an accurate estimate of this derivative.  The
only disadvantage of this technique is the necessity of requiring the functions
to accept complex numbers as arguments. Appendix~\ref{sec:appendixA} includes a
simple example that illustrates the use of the complex-step method for the
derivative calculation of a function. 
%
\note[Michael]{re: II.6}
For the reader's convenience the forward difference equivalent to~\ref{eqn:complexFirstDeriv} is
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x)}{h} +\mathcal{O}\left( h \right). \label{eqn:forwardFirstDeriv} \end{equation},
%
while the central difference equivalent to~\ref{eqn:complexFirstDeriv} is,
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x - h)}{2h} +\mathcal{O}\left( h^2 \right). \label{eqn:centeredFirstDeriv} \end{equation} \cite[Chap. 4.1.3]{chapra2010}.
%
Note that the order of the truncation error in~\ref{eqn:forwardFirstDeriv} indicates that for a given $h$ forward difference is less accurate an approximation than central difference or complex step.

@@

\subsubsection{Automatic differentiation} 
\label{ADsubsection}

Automatic differentiation (AD) is a computerized method for computing exact
derivatives based on the chain-rule from calculus. AD takes advantage of the fact
that any mathematical function executed on a computer, no matter how
complicated, is a ``composition of simple operations'' (addition, multiplication, exponentiation and the like) each having known analytical derivatives
\cite{ref-sacado-presentation}. For reference, the AD implementation used in
this study is part of the \emph{Sacado} package from the \emph{Trilinos}
agile-component libraries developed at Sandia National Laboratories
\cite{ref-Sacado}.

An AD system evaluates composition functions in the expected manner: it works
by first evaluating the innermost function of the composition, then presenting
that function's output as input to the next level function until all levels are
complete, observing commonly expected order of operations. However, an AD
system does additional work during a function evaluation in that as each nested
function is evaluated, the function's partial derivative with respect to the
designated variables of the given input is also calculated.  This is possible
because the elementary math functions are hard-coded into the AD source-code
along with their analytical derivatives, and linked by special instructions, so
that when the elementary math functions are called upon for computation, their
partial derivatives may be computed and stored in a sequence. The AD system
then multiplies the final sequence of partial derivatives together to produce
the exact equivalent to taking a partial derivative of the corresponding
composition function with respect to a designated variable at a particular
value. 
%
It is obvious, but bears mentioning, that the AD system could simply
store one value for partial derivatives, modifying it as appropriate for every
function evaluation rather than keeping a sequence. This is significant because
modifying a single value rather than keeping a list of values corresponding to
every level of a composition function, which itself may not be the sole one
needing to be evaluated, represents a savings in memory usage, which at large
scales is an active concern. In the literature, the AD scheme described here is
called \emph{forward automatic differentiation}. Only forward AD will be
covered here since it is the implementation used in \emph{Sacado} and therefore
in this study; however, the reader is referred to the introduction section of
\cite{ref-AD-methods} and its citations for further information on AD, and
particularly \cite{ref-on-AD} which is foundational.
Appendix~\ref{sec:appendixB} includes an example that walks the reader through
the process a computer uses to compute derivatives via AD.  Some things to note
about AD are that no approximation of derivatives is being made because the
analytical forms of the partial derivatives of the elementary math functions are defined
alongside them. The accuracy of AD is then limited by the precision of the AD
system's definition of the elementary math functions and their partial
derivatives.

\subsection{Tangent-stiffness} 
\label{subsec:TS}

In solid and structural computational mechanics the tangent-stiffness matrix is
a linearization operator that describes the stiffness of a system in response
to small displacements imposed upon the current configuration of the system.
Mathematically speaking, the tangent-stiffness represents the gradient of a
high-dimensional energy surface that ``points'' in the minimum direction. While
the terminology of solid mechanics is used here, it's important to note that
these operators appear in other physical settings and are known by other names,
e.g. a \emph{transmissibility matrix} in the context of a Poisson problem or,
more generally, a \emph{Jacobian matrix} in mathematical optimization.

While, for simplicity, the examples shown in the following refer only to linear
problems, tangent-stiffness matrices generally arise in the context of
non-linear analysis where Newton's method or quasi-Newton methods are used
in the minimization of a given residual function.  In this context, the
tangent-stiffness matrix can be thought of as the linearization of the system
about a particular configuration.  It has been shown
\cite{hughes1978consistent,hughes1978unconditionally} that in order to preserve
the quadratic convergence properties of the Newton's methods, it is necessary
that this linearization is carried out in a manner that is \emph{consistent}
with the algorithmic constraint equations used when computing the internal
forces that arise due to deformations.  An example of these constraint
equations would be the Kuhn-Tucker conditions \cite{simo1998} that are used in
integration of a flow rule for plasticity modeling.  If the continuum
tangent-moduli are used in place of the \emph{consistent} or \emph{algorithmic
tangent moduli} in the solution of a non-linear solid mechanics problem,
convergence may still be achieved, but not in the quadratic manner that makes
Newton's method attractive for this class of problems.  In the setting of
non-linear solid mechanics, there is a very small class of material model
algorithms in which consistent tangent-stiffness operators can be derived
analytically.  Perfect plasticity and plasticity with isotropic hardening used
in combination with a general nearest-point projection or \emph{radial return
algorithm} are a few examples.  Therefore, when general models (and certainly
more complex ones) are implemented into a general purpose computational
mechanics code, the tangent-stiffness operators are typically defined via a
numerical approximation.

A general mathematical formula for a tangent-stiffness matrix, $K_{ij}$, can be
expressed in indicial notation as 
%
\begin{equation} K_{ij} = \left. \frac{\partial F_i}{\partial
X_j}\right|_{\vec{X}_0}, \end{equation}
%
where $F_i$ is the $i^{\mbox{th}}$ component of the vector valued function,
$X_j$ is the $j^{\mbox{th}}$ component of the vector argument of $F$, and a
particular value of the vector, $\vec{X}_0$ is chosen as the linearization
point. One then evaluates the expression for each combination $(i, j)$
corresponding to a (row, column) location in the tangent-stiffness matrix. The
elements of a tangent-stiffness matrix can be estimated using any of the
complex-step, AD, or common finite difference techniques for functions $F:R^1
\rightarrow R^1$, since taking partial derivatives entails holding all but a
single independent vector component of the argument of the function constant, and each element
of the function can be evaluated independently of the others. \\

\subsubsection{Differencing formulas for tangent-stiffness matrix evaluation}
